Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-04-13 writeups/Su_IGW_break.tex
added daniel appendix

The Euler equations for an incompressible fluid in a uniform gravitational field
are\footnote{\textcolor{Corr}{These equations arise by enforcing fluid
incompressibility ($\mdil{\rho}{t} = 0$ or the adiabatic index $\gamma \to
\infty$) in the Euler equations. They also can be derived from the commonly used
``sound-proof'' \emph{anelastic equations} \citep[see e.g.][]{ogura1962scale,
sutherland1} in the limit where (i) the density stratification is weak ($H \gg
L$ for perturbation wavelength $L$) and (ii) the fluid is incompressible (note
the anelastic equations are defined for arbitrary $\gamma$). Under condition
(i), the anelastic approximation reduces to the Boussinesq approximation
\citep{klein2009asymptotics}. Our equations permit a weak but nonzero density
stratification and can be interpreted as ``locally Boussinesq''. Condition (ii)
is motivated by WD fluid properties and, more generally, simplicity, as the
dynamical effect of wave breaking is likely to be similar for different sets of
fluid equations.}}
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $(-g\uv{z})$.

@@

>>>>

The Euler equations for an incompressible fluid in a uniform gravitational field
are
% \footnote{\textcolor{Corr}{These equations arise by enforcing fluid
% incompressibility ($\mdil{\rho}{t} = 0$ or the adiabatic index $\gamma \to
% \infty$) in the Euler equations. They also can be derived from the commonly used
% ``sound-proof'' \emph{anelastic equations} \citep[see e.g.][]{ogura1962scale,
% sutherland1} in the limit where (i) the density stratification is weak ($H \gg
% L$ for perturbation wavelength $L$) and (ii) the fluid is incompressible (note
% the anelastic equations are defined for arbitrary $\gamma$). Under condition
% (i), the anelastic approximation reduces to the Boussinesq approximation
% \citep{klein2009asymptotics}. Our equations permit a weak but nonzero density
% stratification and can be interpreted as ``locally Boussinesq''. Condition (ii)
% filters out sound waves. The dynamical effect of wave breaking is likely to be
% similar for different sets of fluid equations.}}
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $(-g\uv{z})$. \textcolor{Corr}{Note that these equations
conserve the same wave energy as the commonly used anelastic equations
\citep{ogura1962scale,anel_part1} and thus give the same wave amplitude
growth. Appendix~\ref{s:equation_deriv} provides a derivation of these
equations and justification for using them.}

@@

\section{Derivation of Fluid Equations}\label{s:equation_deriv}

We aim to model wave dynamics over multiple density and pressure scaleheights.
Start with the compressible Euler equations.
\begin{align}
    \partial_t \rho + \p{\bm{u}\cdot\bm{\nabla}} \rho
            + \rho \bm{\nabla}\cdot\bm{u} =\, &0, \\
    \partial_t s + \p{\bm{u}\cdot\bm{\nabla}} s =\, &0, \\
    \partial_t \bm{u} + \p{\bm{u}\cdot\bm{\nabla}}\bm{u} +
        \frac{1}{\rho}\bm{\nabla} p =& \, \bm{g},
\end{align}
where $s$ is the entropy. The pressure is calculated from the entropy and
density using the equation of state
\begin{equation}
    \frac{s}{c_p} = \log \frac{p^{1/\gamma}}{\rho},
\end{equation}
where $c_p$ is the specific heat at constant pressure, and $\gamma$ is the ratio
of specific heats. For computational ease, it is convenient to filter the fast
sound waves from these equations. One approach is to assume pressure
perturbations are small \citep[yielding the ``pseudo-incompressible''
equations,][]{anel_part2}, or that all thermodynamic perturbations are small
\citep[yielding the ``anelastic'' equations,][]{anel_part1}. In these
approximations, one of the thermodynamic equations is replaced by a constraint
equation: $\bm{\nabla}\cdot \p{p_0^{1/\gamma} \bm{u}}=0$ for
pseudo-incompressible; $\bm{\nabla}\cdot\p{\rho_0 \bm{u}} =0$ for anelastic.
Here $p_0$ and $\rho_0$ are the background density and pressure profiles. The
pressure in the momentum equation can be interpreted as a Lagrange multiplier
which enforces the constraint \citep{anel_part2}. Upon linearization, both
approximations conserve a wave energy
\begin{equation}
    E_w = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\frac{g^2}{N^2} \frac{(\rho')^2}{\rho_0}
        = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\rho_0 N^2 \frac{(s')^2}{\abs{\bm{\nabla} s_0}^2},
            \label{eq:ewave}
\end{equation}
where $\rho'$ and $s'$ represent the density and entropy perturbations.

Rather than assume thermodynamic perturbations are small, we instead filter
sound waves by taking the limit $\gamma\rightarrow \infty$. Then the entropy and
log density are proportional to each other, and the entropy equation becomes
\begin{equation}
    \partial_t \log \rho + \p{\bm{u}\cdot \bm{\nabla}} \log \rho = 0.
\end{equation}
Together with mass conservation, this implies
\begin{equation}
    \bm{\nabla} \cdot \bm{u} = 0.
\end{equation}
We solve these equations together with the normal momentum equation.

Although these equations are non-standard, they have various desirable
properties. They conserve mass and momentum, and the linearized equations
conserve the wave energy $E_w$ [Eq.~\eqref{eq:ewave} above] similar to the
pseudo-incompressible equations and anelastic equations. Our equations also
satisfy the same linear dispersion relation as the fully compressible equations
in the limit of large sound speed \citep[this is also true for the anelastic
equations, but not pseudo-incompressible,][]{anel_part1, anel_part2}. Thus, the
vertical propagation of internal gravity waves is similar to the
pseudo-incompressible equations and anelastic equations.

In this work we are interested in waves which reach large amplitudes and break.
For breaking waves, \citet{achatz_igw_pi} suggests that the anelastic equations
may miss important effects. Although the pseudo-incompressible equations may
capture wave-breaking more accurately, they are more complicated, and do not
satisfy the correct dispersion relation to order $(k_z H)^{-2}$
\citep{anel_part2}. In the absence of a clear choice to study this wave breaking
problem, we have elected to use these simple equations derived in the
$\gamma\rightarrow\infty$ limit.
